---
title: "Under reported Covid-19 count analysis"
author: " "
footer: "Nathen Byford"
format: baylor-revealjs
view-distance: 5
---

```{r}
#| include: false

library(tidyverse); theme_set(theme_bw())
library(sp)
library(geoR)
library(gstat)
library(plotly)

load("covid.RData")
```

## Covid Count Data

Some information about covid data and why it matters.


## Big Idea

Covid-19 count data is likely under reported due to many reasons:

- It can manifest in different severity levels,
- Some people are apprehensive to get tested,
- Some places don't have the ability to test everyone, and
- etc.

:::{.def} 
Under reported data can bias the estimates to be lower than they really are, can be thought of as unintentional missing data.
:::

## EDA

```{r}
#| include: false
US <- map_data("state")

data <- dat

dat <- dat |> 
  mutate(State = str_to_lower(State))
  
pos_count <- dat |> select(State, positive, Pop) |> rename(region = State)

map_counts <- left_join(US, pos_count)
```

```{r}

data$hover <- with(data, paste(State, '<br>', "Total tests:", totalTestResults,"<br>",
                           "Air polution:", AirPol, "<br>", 
                           "Obesity:", Obesity, "<br>",
                           "Uninsured:", Uninsured))
# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)

fig <- plot_geo(data, locationmode = 'USA-states')
fig <- fig %>% add_trace(
    z = ~positive, text = ~hover, locations = ~Abbre,
    color = ~positive, colors = 'Blues', reversescale = TRUE
  )
fig <- fig %>% colorbar(title = "Cases")
fig <- fig %>% layout(
    title = 'Covid-19 counts by state<br>(Hover for breakdown)',
    geo = g
  )

fig
```

## Methods

- Non-spatial model
- Spatial models
    - Kriging
    - Something else
- Under reported spatial hierarchical model

## Model comparison

- DIC
- WAIC